\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Process mining implementations routinely assert correctness through runtime
checks, test suites, and informal code review. These mechanisms are necessary
but not sufficient: they can verify that a program produced a result, but they
cannot prove that the result was produced by a lawful process. When a crate
claims that an event log has been admitted against the OCEL~2.0 standard, there
is no compile-time obligation that prevents the caller from treating the
admitted evidence as raw, unvalidated data --- unless the type system itself
encodes that distinction. This is the verification gap that
\texttt{wasm4pm-compat} addresses.

We present a nightly Rust type-law foundry whose central invariant is a
typed, one-way lifecycle enforced entirely at compile time. The carrier type
\Evidence{T}{State}{W} couples a value with a state token and a witness marker
that names the specific paper, standard, or law under which the value was
admitted. Because state tokens are empty enums and witness markers are
zero-sized \texttt{PhantomData} fields, the distinction between
\Evidence{T}{\texttt{Raw}}{W} and \Evidence{T}{\texttt{Admitted}}{W} is
erased at runtime while remaining fully opaque to the type checker. No cast,
coercion, or unsafe block can bridge two incompatible lifecycle states; the
compiler rejects such programs before they execute.

The approach is certified through the ALIVE gate: a trybuild-based
compile-time receipt suite in \texttt{tests/ui/}. The gate consists of
compile-fail fixtures --- each of which must fail for the \emph{intended named
law}, not accidentally --- and compile-pass fixtures that prove every lawful path
remains open. Reaching \textsc{PaperLaw Crown Alive 004} required
37~modules, 196~compile-fail receipts, 406~compile-pass receipts, 98~mapped
papers from the process mining literature, and 601~receipt-bearing commits
manufactured across May~2026.

The principal contribution is the \emph{manufacturing doctrine}: a discipline
in which every commit is a receipt, every compile-fail fixture is a proof
obligation, and the graduation boundary to the \texttt{wasm4pm} execution
engine is enforced by feature-gated type contracts rather than by convention.
The result is a crate that applications cannot misuse at the lifecycle level
without a compiler error, and whose correctness evidence survives intact
across refactors, toolchain upgrades, and contributor changes.

\bigskip
\noindent\textbf{Keywords:} process mining, type theory, nightly Rust,
compile-time verification, object-centric event logs, OCEL~2.0, trybuild,
ALIVE gate, manufacturing doctrine, evidence lifecycle.
